My title

### 2024-09-23 paper/paper.qmd
paper data section updated

# Load packages
package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr")

@@

# Load data

@@

Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], .

@@

To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department. These records are geospatially coded (as seen in the `geometry` column), allowing each cycling path to be mapped precisely in relation to other city infrastructure. Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history. This allows researchers to analyze trends in cycling infrastructure and its expansion over time.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-one
#| tbl-cap: "cap"
#| warning: false

>>>>

# Load packages
package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr", "here", "knitr", "kableExtra")

@@

@@

Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be talked about in the @sec-data-method.

@@

To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department. These records are geospatially coded (column geometry in this dataset), allowing each cycling path to be mapped precisely in relation to other city infrastructure. Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history. This allows researchers to analyze trends in cycling infrastructure and its expansion over time.

@@

Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR]. Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], `here` [], `knitr` [], `kableExtra` [], `` [], `` [], .

```{r}
#| echo: false
#| eval: true
#| label: tbl-raw-coordinates
#| tbl-cap: "Raw Coordinates"
#| warning: false

# Load the dataset
data <- read_csv(here("data", "raw_data", "raw_data.csv"))

# Extract the first part of the geometry string for each row
data_geometry <- data %>%
  mutate(simplified_geometry = substr(geometry, 1, 92)) %>%  # Extract the first 100 characters
  select(simplified_geometry)

# Display the simplified geometry data
head(data_geometry) %>%
  kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  column_spec(1, width = "40em")  # Adjust the width as needed
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-clean-coordinates
#| tbl-cap: "Extracted and Cleaned Coordinates"
#| warning: false

# Load the dataset
data <- read_csv(here("data", "raw_data", "raw_data.csv"))

# Function to extract and clean coordinates from geometry
extract_coordinates <- function(geo) {
  if (!is.na(geo)) {
    geo_clean <- gsub("'", "\"", geo)
    json_data <- tryCatch({
      fromJSON(geo_clean)
    }, error = function(e) {
      return(NULL)
    })

    if (!is.null(json_data) && !is.null(json_data$coordinates)) {
      coords <- unlist(json_data$coordinates, recursive = TRUE)
      longitude <- coords[grepl("^-79", coords)]
      latitude <- coords[grepl("^43", coords)]
      if (length(longitude) == length(latitude)) {
        coords_matrix <- cbind(longitude, latitude)
        return(as.data.frame(coords_matrix))
      }
    }
  }
  return(NULL)
}

# Extract coordinates and clean data
all_coordinates <- lapply(data$geometry, extract_coordinates)
valid_coordinates <- all_coordinates[!sapply(all_coordinates, is.null)]
data$id <- seq_len(nrow(data))
coordinates_data <- do.call(rbind, Map(function(coords, id) {
  coords$id <- id
  return(coords)
}, valid_coordinates, data$id[!sapply(all_coordinates, is.null)]))

# Set column names and remove NA values
colnames(coordinates_data) <- c("Longitude", "Latitude", "ID")
coordinates_data <- coordinates_data %>% filter(!is.na(Longitude), !is.na(Latitude))

# Display the first few rows of the cleaned coordinate data
head(coordinates_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Longitude", "Latitude", "ID"))
```

In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which contains location data in JSON-like format, @tbl-raw-coordinates displays how the data was like. A custom function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values. The extracted coordinates were combined into a new dataset shon in @tbl-clean-coordinates, with unique IDs to match the original entries. Any rows with missing coordinate data were removed, and the cleaned data was saved into a separate file for further analysis.

```{r}
#| echo: false
#| eval: true
#| label: tbl-clean-upgraded
#| tbl-cap: "Number of Lane Upgrades by Year"
#| warning: false

# Load the dataset
data <- read_csv(here("data", "raw_data", "raw_data.csv"))

# Step 1: Select the 'UPGRADED' column
upgraded_data <- data %>%
  select(UPGRADED)

# Step 2: Filter out invalid and missing years
upgraded_data <- upgraded_data %>%
  filter(!is.na(UPGRADED) & UPGRADED > 1800)  # Keep valid years only

# Step 3: Convert the 'UPGRADED' column to numeric
upgraded_data <- upgraded_data %>%
  mutate(UPGRADED = as.numeric(UPGRADED))

# Step 4: Group the data by the year of upgrade and summarize the number of upgrades
upgraded_summary <- upgraded_data %>%
  group_by(UPGRADED) %>%
  summarise(num_lanes = n())

# Step 5: Display the summarized data in a table
tail(upgraded_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))
```

The second cleaning process dealt with the 'UPGRADED' column, which contains information about the years when cycling lanes were upgraded. Invalid or missing years were filtered out, and the remaining valid data was converted to a numeric format. The dataset was then grouped by the year of upgrade, with a summary created to count the number of lanes upgraded in each year. This summarized data, @tbl-clean-upgraded, was saved for use in analyzing trends in lane upgrades over time.

```{r}
#| echo: false
#| eval: true
#| label: tbl-clean-installed
#| tbl-cap: "Number of Bikeways Installed by Year"
#| warning: false

# Load the dataset
data <- read_csv(here("data", "raw_data", "raw_data.csv"))

# Step 1: Select the 'INSTALLED' column
installed_data <- data %>%
  select(INSTALLED)

# Step 2: Filter out invalid and missing years
installed_data <- installed_data %>%
  filter(!is.na(INSTALLED) & INSTALLED > 1800)  # Keep valid years only

# Step 3: Convert the 'INSTALLED' column to numeric
installed_data <- installed_data %>%
  mutate(INSTALLED = as.numeric(INSTALLED))

# Step 4: Group the data by the year of installation and summarize the number of installations
installed_summary <- installed_data %>%
  group_by(INSTALLED) %>%
  summarise(num_bikeways = n())

# Step 5: Display only the first 6 rows of the summarized data in a table
head(installed_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))
```

In the third cleaning process, the focus was on the 'INSTALLED' column, which records the years when cycling infrastructure was first installed. Similar to the 'UPGRADED' column, invalid or missing years were removed, and the data was converted to numeric format. The installation data was grouped by year to count the number of bikeways installed each year. The resulting summary, @tbl-clean-installed, was saved for further trend analysis on the growth of the cycling network.

```{r}
#| echo: false
#| eval: true
#| label: tbl-clean-comfort
#| tbl-cap: "Classification of Cycling Lanes by Comfort Level"
#| warning: false

# Load the dataset
data <- read_csv(here("data", "raw_data", "raw_data.csv"))

# Step 1: Select relevant columns for classification
lane_type_data <- data %>%
  select(INFRA_HIGHORDER, INFRA_LOWORDER, INSTALLED, UPGRADED)

# Step 2: Filter out rows with missing values in the relevant columns
lane_type_data <- lane_type_data %>%
  filter(!is.na(INFRA_HIGHORDER) & !is.na(INFRA_LOWORDER))

# Step 3: Classify lanes into comfort levels based on infrastructure types
lane_type_data <- lane_type_data %>%
  mutate(
    Comfort_Level = case_when(
      str_detect(INFRA_HIGHORDER, "Protected|Multi-Use") | str_detect(INFRA_LOWORDER, "Protected|Multi-Use") ~ "High Comfort",
      str_detect(INFRA_HIGHORDER, "Bike Lane") | str_detect(INFRA_LOWORDER, "Bike Lane") ~ "Moderate Comfort",
      TRUE ~ "Low Comfort"
    )
  )

# Step 4: Select relevant columns to display (Comfort Level, Installed, and Upgraded year)
final_lane_data <- lane_type_data %>%
  select(INSTALLED, UPGRADED, Comfort_Level)

# Step 5: Display the first 6 rows of the cleaned and classified data
head(final_lane_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Installed Year", "Upgraded Year", "Comfort Level"))
```

Finally, the fourth cleaning process involved classifying cycling lanes based on the type of infrastructure. Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed. The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on whether the infrastructure was protected or involved bike lanes. The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels. @tbl-clean-comfort shows how it looks like.
